---
title: "hw1"
author: "TING-YU CHANG"
format: html
editor: visual
---

## Q1:Exploratory Analysis of Soybean Cultivars

### a. Import the data into a data.frame in R. Use the dataset documentation to assign appropriate variable names. (Note: Downloading and unzipping the file can take place outside of your submitted document, but importing the file should be in the submission.)

```{r}
soybean <- read.csv("data.csv")

colnames(soybean) <- c(
  "Season",
  "Cultivar",
  "Repetition",
  "PH",
  "IFP",
  "NLP",
  "NGP",
  "NGL",
  "NS",
  "MHG",
  "GY"
)

```

### b. Using the information provided in the dataset documentation: Verify that there are 320 observations in the dataset; 40 distinct cultivars; and all the observations were collected from two seasons. Produce a table showing the number of observations from each season.

```{r}
# number of obs.
nrow(soybean)

# number of distinct cultivars
length(unique(soybean$Cultivar))

# number of distinct seasons
length(unique(soybean$Season))

# number of observations from each season
table(soybean$Season)
```

### The dataset contains 320 observations, 40 distinct cultivars, and two seasons. Each season contains 160 observations.

### c. Use the data to answer the following questions

### 1.What is the correlation between plant height (PH) and grain yield (GY)?

```{r}
cor(soybean$PH, soybean$GY)
```

### The correlation between plant height (PH) and grain yield (GY) is 0.1232807.

### 2.Calculate the correlation between PH and GY separately for each season. Which season has the higher correlation?

```{r}
cor(soybean$PH[soybean$Season==1],
    soybean$GY[soybean$Season==1])

cor(soybean$PH[soybean$Season==2],
    soybean$GY[soybean$Season==2])
```

### Season 2 has the higher correlation.

### 3.What is the grain yield of the soybean observation with the greatest plant height?

```{r}
soybean[soybean$PH == max(soybean$PH), c("PH","GY")]
```

### The grain yield of the soybean observation with the greatest plant height = 3304.678.

### 4.What percentage of observations have a thousand seed weight (MHG) greater than the overall average thousand seed weight?

```{r}
mean_MHG <- mean(soybean$MHG)
# covert to %
mean(soybean$MHG > mean_MHG) * 100
```

### 47.1875%

### 5.Which cultivar has the highest average grain yield?

```{r}
# calculate average GY
avg_GY <- aggregate(GY ~ Cultivar,
                    data = soybean,
                    FUN = mean)
# find the highest average GY
avg_GY[which.max(avg_GY$GY), ]
```

### The answer is 98R30 CE.

### d. Create a table showing the average value of each quantitative variable: One row containing overall averages. One row for each season containing season-specific averages. The table should clearly identify what each row represents. (This table does not need to be “fancy” but should clearly identify what each value represents.)

```{r}
# quantitative variables
quant_vars <- c("PH","IFP","NGP","NGL","NS","MHG","GY")

# overall averages
overall_avg <- data.frame(
  Group = "Overall",
  t(colMeans(soybean[quant_vars]))
)

# season-specific averages
season_avg <- aggregate(
  soybean[quant_vars],
  by = list(Group = paste0("Season ", soybean$Season)),
  FUN = mean
)

#combine into one table
avg_table <- rbind(overall_avg, season_avg)

avg_table
```
