---
title: "hw1"
author: "TING-YU CHANG"
format: html
editor: visual
---

## Q1:Exploratory Analysis of Soybean Cultivars

### a. Import the data into a data.frame in R. Use the dataset documentation to assign appropriate variable names. (Note: Downloading and unzipping the file can take place outside of your submitted document, but importing the file should be in the submission.)

```{r}
soybean <- read.csv("data.csv")

colnames(soybean) <- c(
  "Season",
  "Cultivar",
  "Repetition",
  "PH",
  "IFP",
  "NLP",
  "NGP",
  "NGL",
  "NS",
  "MHG",
  "GY"
)

```

### b. Using the information provided in the dataset documentation: Verify that there are 320 observations in the dataset; 40 distinct cultivars; and all the observations were collected from two seasons. Produce a table showing the number of observations from each season.
```{r}
# number of obs.
nrow(soybean)

# number of distinct cultivars
length(unique(soybean$Cultivar))

# number of distinct seasons
length(unique(soybean$Season))

# number of observations from each season
table(soybean$Season)
```
### The dataset contains 320 observations, 40 distinct cultivars, and two seasons. Each season contains 160 observations.

### c. Use the data to answer the following questions:

1.What is the correlation between plant height (PH) and grain yield (GY)?
2.Calculate the correlation between PH and GY separately for each season. Which season has the higher correlation?
3.What is the grain yield of the soybean observation with the greatest plant height?
4.What percentage of observations have a thousand seed weight (MHG) greater than the overall average thousand seed weight?
5.Which cultivar has the highest average grain yield?


