---
title: "hw1"
author: "TING-YU CHANG"
format: html
editor: visual
---

## Q1:Exploratory Analysis of Soybean Cultivars

### a. Import the data into a data.frame in R. Use the dataset documentation to assign appropriate variable names. (Note: Downloading and unzipping the file can take place outside of your submitted document, but importing the file should be in the submission.)

```{r}
soybean <- read.csv("data.csv")

colnames(soybean) <- c(
  "Season",
  "Cultivar",
  "Repetition",
  "PH",
  "IFP",
  "NLP",
  "NGP",
  "NGL",
  "NS",
  "MHG",
  "GY"
)

```

### b. Using the information provided in the dataset documentation: Verify that there are 320 observations in the dataset; 40 distinct cultivars; and all the observations were collected from two seasons. Produce a table showing the number of observations from each season.

```{r}
# number of obs.
nrow(soybean)

# number of distinct cultivars
length(unique(soybean$Cultivar))

# number of distinct seasons
length(unique(soybean$Season))

# number of observations from each season
table(soybean$Season)
```

The dataset contains 320 observations, 40 distinct cultivars, and two seasons. Each season contains 160 observations.

### c. Use the data to answer the following questions

### 1.What is the correlation between plant height (PH) and grain yield (GY)?

```{r}
cor(soybean$PH, soybean$GY)
```

The correlation between plant height (PH) and grain yield (GY) is 0.1232807.

### 2.Calculate the correlation between PH and GY separately for each season. Which season has the higher correlation?

```{r}
cor(soybean$PH[soybean$Season==1],
    soybean$GY[soybean$Season==1])

cor(soybean$PH[soybean$Season==2],
    soybean$GY[soybean$Season==2])
```

Season 2 has the higher correlation.

### 3.What is the grain yield of the soybean observation with the greatest plant height?

```{r}
soybean[soybean$PH == max(soybean$PH), c("PH","GY")]
```

The grain yield of the soybean observation with the greatest plant height = 3304.678.

### 4.What percentage of observations have a thousand seed weight (MHG) greater than the overall average thousand seed weight?

```{r}
mean_MHG <- mean(soybean$MHG)
# covert to %
mean(soybean$MHG > mean_MHG) * 100
```

47.1875%

### 5.Which cultivar has the highest average grain yield?

```{r}
# calculate average GY
avg_GY <- aggregate(GY ~ Cultivar,
                    data = soybean,
                    FUN = mean)
# find the highest average GY
avg_GY[which.max(avg_GY$GY), ]
```

The answer is 98R30 CE.

### d. Create a table showing the average value of each quantitative variable: One row containing overall averages. One row for each season containing season-specific averages. The table should clearly identify what each row represents. (This table does not need to be “fancy” but should clearly identify what each value represents.)

```{r}
# quantitative variables
quant_vars <- c("PH","IFP","NGP","NGL","NS","MHG","GY")

# overall averages
overall_avg <- data.frame(
  Group = "Overall",
  t(colMeans(soybean[quant_vars]))
)

# season-specific averages
season_avg <- aggregate(
  soybean[quant_vars],
  by = list(Group = paste0("Season ", soybean$Season)),
  FUN = mean
)

#combine into one table
avg_table <- rbind(overall_avg, season_avg)

avg_table
```

### e. Carry out a series of t-tests to examine whether the ngrain yield (GY) differs between the two seasons. State the null and the alternative hypotheses. Present the R output and interpret the results.

```{r}
# calculate GY by season
GY1 <- soybean$GY[soybean$Season == 1]
GY2 <- soybean$GY[soybean$Season == 2]

# sample statistics
n1 <- length(GY1)
n2 <- length(GY2)

mean1 <- mean(GY1)
mean2 <- mean(GY2)

var1 <- var(GY1)
var2 <- var(GY2)

# calculate t statistics
t_stat <- (mean1 - mean2) / sqrt(var1 / n1 + var2 / n2)

# calculate df
df <- (var1 / n1 + var2 / n2)^2 / ((var1 / n1)^2 / (n1 - 1) + (var2 / n2)^2 / (n2 - 1))

# calculate two-sided p-value
p_value <- 2 * pt(-abs(t_stat), df = df)

t_stat
df
p_value

t.test(GY ~ Season, data = soybean)
```

Null hypothesis: $H_0: \mu_1 = \mu_2$, meaning that the mean grain yield is the same in the two seasons. Alternative hypothesis: $H_0: \mu_1 neq \mu_2$, meaning that the mean grain yield differs between the two seasons. The Welch two-sample t-test gave (t = 0.35155), (df = 231.85), and a p-value of 0.7255. Since the p-value is greater than 0.05, we fail to reject the null hypothesis. There is not sufficient evidence to conclude that the mean grain yield differs between the two seasons.
