---
title: "hw1"
author: "TING-YU CHANG"
format: html
editor: visual
---

## Q1:Exploratory Analysis of Soybean Cultivars

### a. Import the data into a data.frame in R. Use the dataset documentation to assign appropriate variable names. (Note: Downloading and unzipping the file can take place outside of your submitted document, but importing the file should be in the submission.)

```{r}
soybean <- read.csv("data.csv")

colnames(soybean) <- c(
  "Season",
  "Cultivar",
  "Repetition",
  "PH",
  "IFP",
  "NLP",
  "NGP",
  "NGL",
  "NS",
  "MHG",
  "GY"
)

```

### b. Using the information provided in the dataset documentation: Verify that there are 320 observations in the dataset; 40 distinct cultivars; and all the observations were collected from two seasons. Produce a table showing the number of observations from each season.
```{r}
# number of obs.
nrow(soybean)

# number of distinct cultivars
length(unique(soybean$Cultivar))

# number of distinct seasons
length(unique(soybean$Season))

# number of observations from each season
table(soybean$Season)
```
### The dataset contains 320 observations, 40 distinct cultivars, and two seasons. Each season contains 160 observations.

###


