---
title: "Datasci506_hw2"
author: "TING-YU CHANG"
format: html
editor: visual
---

This is the link to my repo on github:https://github.com/tingyuuucc/stats506-HW1

## Problem 1: Random Walk with Momentum

### a. Simulation function:
```{r}
momentum_walk <- function(n, p) {
  # check n is a single positive integer
  if(length(n) != 1 || is.na(n) || !is.numeric(n) ||
     n <= 0 || n != floor(n)) {
    stop("n must be a positive integer.")
  }
  
  # check p is a single number between 0 and 1
  if (length(p) != 1 || is.na(p) || !is.numeric(p) ||
      p < 0 || p > 1) {
    stop("p must be a number between 0 and 1.")
  }
  
  # define 4 posiible directions
  directions <- matrix(
    c(1, 0,    # right
      -1, 0,   # left
      0, 1,    # up
      0, -1),  # down
    ncol = 2,
    byrow = TRUE
  )
  # start at origin
  position <- c(0,0)
  
  # choose 1st direction
  pre_direction <- sample(1:4, 1)
  
  # first step
  position <- position + directions[pre_direction, ]
  
  # remaining steps
  if (n > 1) {
    for (i in 2:n) {
      if (runif(1) < p) {
        current_direction <- pre_direction
      } else {
        current_direction <- sample(1:4, 1)
      }
      position <- position + directions[current_direction, ]
      pre_direction <- current_direction
    }
  }
  return(position)
}
```

```{r}
momentum_walk(10, 0.6)
momentum_walk(10, 0.6)
```

### b. Implement the simulation

### version 1
```{r}
momentum_walk1 <- function(n, p) {
  if (length(n) != 1 || is.na(n) || !is.numeric(n) ||
      n <= 0 || n != floor(n)) {
    stop("n must be a positive integer.")
  }
  
  if (length(p) != 1 || is.na(p) || !is.numeric(p) ||
      p < 0 || p > 1) {
    stop("p must be a number between 0 and 1.")
  }
  
  directions <- matrix(
    c(1, 0,
     -1, 0,
      0, 1,
      0, -1),
    ncol = 2,
    byrow = TRUE
  )
  
  position <- c(0, 0)
  
  prev_direction <- sample(1:4, 1)
  position <- position + directions[prev_direction, ]
  
  if (n > 1) {
    for (i in 2:n) {
      
      # preserve the previous direction with probability p
      if (runif(1) < p) {
        current_direction <- prev_direction
      } else {
        current_direction <- sample(1:4, 1)
      }
      position <- position + directions[current_direction, ]
      prev_direction <- current_direction
    }
  }
  
  return(position)
}
```

### version 2

```{r}
momentum_walk2 <- function(n, p) {
  
  if (length(n) != 1 || is.na(n) || !is.numeric(n) ||
      n <= 0 || n != floor(n)) {
    stop("n must be a positive integer.")
  }

  if (length(p) != 1 || is.na(p) || !is.numeric(p) ||
      p < 0 || p > 1) {
    stop("p must be a number between 0 and 1.")
  }
  
  directions <- matrix(
    c(1, 0,
     -1, 0,
      0, 1,
      0, -1),
    ncol = 2,
    byrow = TRUE
  )
  
  first_direction <- sample(1:4, 1)
  previous_direction <- first_direction
  
  if (n > 1) {
    
    remaining_directions <- sapply(seq_len(n - 1), function(i) {
      
      if (runif(1) < p) {
        current_direction <- previous_direction
      } else {
        current_direction <- sample(1:4, 1)
      }
      
      previous_direction <<- current_direction
      
      return(current_direction)
    })
    
    # combine all step directions
    step_directions <- c(first_direction, remaining_directions)
    
  } else {
    step_directions <- first_direction
  }
  
  # add all step to obtain the final position
  final_position <- colSums(directions[step_directions, , drop = FALSE])
  
  return(final_position)
}
```

```{r}
# demonstrate the loop version
momentum_walk1(10, 0.6)

# demonstrate the apply version
momentum_walk2(10, 0.6)

# demonstrate both functions with 1000 steps
momentum_walk1(1000, 0.6)
momentum_walk2(1000, 0.6)
```
### momentum_walk1() uses a for loop to generate each step and update the position. momentum_walk2() uses sapply() to generate the sequence of directions, then sums the corresponding movements to obtain the final position.

### ChatGPT helped me figure out a more briefly way to show my result.







